Effectful Semantics for Substructural SSA

### 2023-11-09 pldi24/paper.tex
Tap

@@

\TODO{late \ms{where}-binding, and other sugar (?)}

\TODO{top-level functions}

\TODO{table of typing judgements}

\TODO{contexts, label-contexts}

@@

>>>>

In Figure \ref{fig:ssa-grammar}, we present a formal grammar for our SSA syntax. We begin by giving a grammar for \textit{types} \(A, B, C\), parametrized by a set of \textit{base types} \(X\). We require the existence of a \textit{unit type} \(\mb{1}\), a \textit{boolean type} \(\mb{2}\), and all tensor products \(A \otimes B\); we write \(\otimes\) to underscore the fact that types are allowed to be substructural.

We separate program fragments into 3 main syntactic categories: \textit{expressions} \(a, b, c, e\), \textit{blocks} \(s, t\), and \textit{label-sets} \(L\). While we make a distinction between \textit{pure} and \textit{impure} expressions, this is done at the level of typing rules rather than via syntax. A \textit{block} is conceptualized as a series of \textit{let-bindings} followed by a \textit{terminator}. Here, a terminator is either an unconditional branch to another block with an argument or an if-statement. A block may be encapsulated within a \textit{where-block}, which defines other blocks the inner block can branch to; these blocks can also branch to each other in a mutually recursive fashion. Similarly to blocks, expressions may also contain let-bindings; this is somewhat of a departure from classical SSA, in which there is only one kind of variable binding, but facilitates various optimizations and corresponds more closely to the in-memory representation, where pure expressions are represented as E-graph nodes. Note that we do \textit{not} give a grammar or semantics for top-level functions, which we can view as simply a list of arguments and a block. Except for inlining, we will only consider the optimization and analysis of a single function body at a time.

We proceed to give a grammar for \textit{contexts} \(\Gamma\), which consist of a list of variables \(x\) given a type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). This tells us whether the variable is \textit{affine} (i.e., can be ignored) and/or \textit{relevant} (i.e., whether it can be used multiple times). A variable which is both affine and relevant is called \textit{intuitionistic}. Whether we can use a variable also depends on the \textit{linearity} of its type.

We can then give a grammar for \textit{label-contexts} \(\ms{L}\), which consist of a list of \textit{labels} each annotated with a context \(\Gamma\), denoting the variables which must be live on entry to the block (corresponding to the set a dominance frontier would compute), and an \textit{argument} \(A\).

@@

@@

We can now list out our primary typing judgements, which are given in Figure \ref{fig:ssa-judgements}.
